---
title: "Lab 8"
author: "Antonio Jurlina"
date: "11/14/2020"
output: pdf_document
fontsize: 10pt 
geometry: margin=1in
urlcolor: blue
---

```{r setup, include=FALSE}
#-------- packages --------
library(tidyverse)
library(ggthemes)
library(gstat)
library(sp)
library(sf)
library(gridExtra)
library(knitr)

#-------- data --------
xy <- expand.grid(1:100, 1:100)
names(xy) <- c('x','y')

#-------- models --------
#Model_1 spatially random
formula1=z~1
c_model1=vgm(psill=1.0, range=0, model='Nug') # random covariance model
beta1=1

model1 <- gstat(formula=formula1, locations=~x+y, dummy=T, beta=beta1, model=c_model1, nmax=20)
ff1 <- predict(model1, newdata=xy, nsim=1)
simd1 <- ff1
gridded(simd1) = ~x+y

plot1 <- ggplot()+
	geom_point(data=ff1,aes(x=x,y=y,col=sim1)) +
	scale_colour_gradient(low="blue",high="green") +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Spatially random") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

#Model_2 linear trend only in x direction
formula2=z~x+y
c_model2=vgm(psill=1.0, range=0, model='Nug')
beta2=c(1,-0.1,0.00) #trend coefficients

model2 <- gstat(formula=formula2, locations=~x+y, dummy=T, beta=beta2, model=c_model2, nmax=20)
ff2 <- predict(model2, newdata=xy, nsim=1)
simd2 <- ff2
gridded(simd2) = ~x+y

plot2 <- ggplot()+
	geom_point(data=ff2,aes(x=x,y=y,col=sim1)) +
	scale_colour_gradient(low="blue",high="green") +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Linear trend in x") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

#Model_3 Gaussian spatial dependence only
formula3=z~1
c_model3=vgm(psill=.75, range=15, nugget=.25, model='Gau')
beta3=1

model3 <- gstat(formula=formula3, locations=~x+y, dummy=T, beta=beta3, model=c_model3, nmax=20)
ff3 <- predict(model3, newdata=xy, nsim=1)
simd3 <- ff3
gridded(simd3) = ~x+y

plot3 <- ggplot()+
	geom_point(data=ff3,aes(x=x,y=y,col=sim1)) +
	scale_colour_gradient(low="blue",high="green") +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Gaussian spatial dependence") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

#Model_4 linear trend plus Gaussian spatial dependence
formula4=z~x+y
c_model4=vgm(psill=.75, range=15, nugget=.25, model='Gau')
beta4=c(1,-0.1,0.00)

model4 <- gstat(formula=formula4, locations=~x+y, dummy=T, beta=beta4, model=c_model4, nmax=20)
ff4 <- predict(model4, newdata=xy, nsim=1)
simd4 <- ff4
gridded(simd4) = ~x+y

plot4 <- ggplot()+
	geom_point(data=ff4,aes(x=x,y=y,col=sim1)) +
	scale_colour_gradient(low="blue",high="green") +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Gausian spatial dependence + linear trend") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

models <- list(model1, model2, model3, model4)

#-------- Spatially random model sampling --------
sptest1 <- spsample(simd1, 50, type = "random")
spdf1 <- as.data.frame(sptest1)

plot1.1 <- ggplot()+
	geom_point(data=ff1,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf1, aes(x=x, y=y), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Simple random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest1 <- spsample(simd1, 50, type = "stratified")
spdf1 <- as.data.frame(sptest1)

plot2.1 <- ggplot()+
	geom_point(data=ff1,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf1, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Stratified random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest1 <- spsample(simd1, 50, type = "regular")
spdf1 <- as.data.frame(sptest1)

plot3.1 <- ggplot()+
	geom_point(data=ff1,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf1, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Systematic sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

#-------- First order trend model sampling --------
sptest2 <- spsample(simd2, 50, type = "random")
spdf2 <- as.data.frame(sptest2)

plot1.2 <- ggplot()+
	geom_point(data=ff2,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf2, aes(x=x, y=y), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Simple random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest2 <- spsample(simd2, 50, type = "stratified")
spdf2 <- as.data.frame(sptest2)

plot2.2 <- ggplot()+
	geom_point(data=ff2,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf2, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Stratified random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest2 <- spsample(simd2, 50, type = "regular")
spdf2 <- as.data.frame(sptest2)

plot3.2 <- ggplot()+
	geom_point(data=ff2,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf2, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Systematic sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

#-------- Second order trend model sampling --------
sptest3 <- spsample(simd3, 50, type = "random")
spdf3 <- as.data.frame(sptest3)

plot1.3 <- ggplot()+
	geom_point(data=ff3,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf3, aes(x=x, y=y), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Simple random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest3 <- spsample(simd3, 50, type = "stratified")
spdf3 <- as.data.frame(sptest3)

plot2.3 <- ggplot()+
	geom_point(data=ff3,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf3, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Stratified random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest3 <- spsample(simd3, 50, type = "regular")
spdf3 <- as.data.frame(sptest3)

plot3.3 <- ggplot()+
	geom_point(data=ff3,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf3, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Systematic sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

#-------- First and second order trend model sampling --------
sptest4 <- spsample(simd4, 50, type = "random")
spdf4 <- as.data.frame(sptest4)

plot1.4 <- ggplot()+
	geom_point(data=ff4,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf4, aes(x=x, y=y), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Simple random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest4 <- spsample(simd4, 50, type = "stratified")
spdf4 <- as.data.frame(sptest4)

plot2.4 <- ggplot()+
	geom_point(data=ff4,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf3, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Stratified random sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))

sptest4 <- spsample(simd4, 50, type = "regular")
spdf4 <- as.data.frame(sptest4)

plot3.4 <- ggplot()+
	geom_point(data=ff4,aes(x=x,y=y,col=sim1), alpha = 0.7, show.legend = F) +
	scale_colour_gradient(low="blue",high="green") +
	geom_point(data=spdf4, aes(x=x1, y=x2), size = 2) +
	coord_fixed(ratio=1) +
	theme_minimal() +
	labs(subtitle = "Systematic sampling") +
	theme(axis.title.x = element_blank(),
				axis.title.y = element_blank(),
				plot.margin=unit(c(0,0,0,0),"pt"))
```

```{r functions, include=FALSE}
closest_point <- function(sample.pt, grid.data){
	dist.sq <- (coordinates(grid.data)[,1] - sample.pt[1])^2 +
		(coordinates(grid.data)[,2] - sample.pt[2])^2
	return(which.min(dist.sq))
}

rand_samp <- function (model) {
	simd <- predict(model, newdata=xy, nsim=1)
	gridded(simd) = ~x+y

	# Create the locations of the random sample
	spsamp.pts <- spsample(simd, 50, type = "random")
	# Extract a two column array of the x and y coords
	sample.coords <- coordinates(spsamp.pts)
	# Apply the function closest.point() to each row of the
	# array sample.coords (i.e., each sample location)
	samp.pts <- apply(sample.coords, 1, closest_point,
										grid.data = simd1)
	# Each element of samp.pts is the index of the population value
	# closest to the corresponding location in sample.coords
	data.samp <- simd1$sim1[samp.pts]
	samp.mean.50.random <- mean(data.samp)
	prct.err.50.random <- 100 * abs(mean(data.samp) - mean(simd1$sim1)) / mean(simd1$sim1)

	# Create the locations of the random sample
	spsamp.pts <- spsample(simd, 150, type = "random")
	# Extract a two column array of the x and y coords
	sample.coords <- coordinates(spsamp.pts)
	# Apply the function closest.point() to each row of the
	# array sample.coords (i.e., each sample location)
	samp.pts <- apply(sample.coords, 1, closest_point,
										grid.data = simd1)
	# Each element of samp.pts is the index of the population value
	# closest to the corresponding location in sample.coords
	data.samp <- simd1$sim1[samp.pts]
	samp.mean.150.random <- mean(data.samp)
	prct.err.150.random <- 100 * abs(mean(data.samp) - mean(simd1$sim1)) / mean(simd1$sim1)

	# Create the locations of the stratified sample
	spsamp.pts <- spsample(simd, 50, type = "stratified")
	# Extract a two column array of the x and y coords
	sample.coords <- coordinates(spsamp.pts)
	# Apply the function closest.point() to each row of the
	# array sample.coords (i.e., each sample location)
	samp.pts <- apply(sample.coords, 1, closest_point,
										grid.data = simd1)
	# Each element of samp.pts is the index of the population value
	# closest to the corresponding location in sample.coords
	data.samp <- simd1$sim1[samp.pts]
	samp.mean.50.stratified <- mean(data.samp)
	prct.err.50.stratified <- 100 * abs(mean(data.samp) - mean(simd1$sim1)) / mean(simd1$sim1)

	# Create the locations of the stratified sample
	spsamp.pts <- spsample(simd, 150, type = "stratified")
	# Extract a two column array of the x and y coords
	sample.coords <- coordinates(spsamp.pts)
	# Apply the function closest.point() to each row of the
	# array sample.coords (i.e., each sample location)
	samp.pts <- apply(sample.coords, 1, closest_point,
										grid.data = simd1)
	# Each element of samp.pts is the index of the population value
	# closest to the corresponding location in sample.coords
	data.samp <- simd1$sim1[samp.pts]
	samp.mean.150.stratified <- mean(data.samp)
	prct.err.150.stratified <- 100 * abs(mean(data.samp) - mean(simd1$sim1)) / mean(simd1$sim1)

	# Create the locations of the regular sample
	spsamp.pts <- spsample(simd, 50, type = "regular")
	# Extract a two column array of the x and y coords
	sample.coords <- coordinates(spsamp.pts)
	# Apply the function closest.point() to each row of the
	# array sample.coords (i.e., each sample location)
	samp.pts <- apply(sample.coords, 1, closest_point,
										grid.data = simd1)
	# Each element of samp.pts is the index of the population value
	# closest to the corresponding location in sample.coords
	data.samp <- simd1$sim1[samp.pts]
	samp.mean.50.regular <- mean(data.samp)
	prct.err.50.regular <- 100 * abs(mean(data.samp) - mean(simd1$sim1)) / mean(simd1$sim1)

	# Create the locations of the regular sample
	spsamp.pts <- spsample(simd, 150, type = "regular")
	# Extract a two column array of the x and y coords
	sample.coords <- coordinates(spsamp.pts)
	# Apply the function closest.point() to each row of the
	# array sample.coords (i.e., each sample location)
	samp.pts <- apply(sample.coords, 1, closest_point,
										grid.data = simd1)
	# Each element of samp.pts is the index of the population value
	# closest to the corresponding location in sample.coords
	data.samp <- simd1$sim1[samp.pts]
	samp.mean.150.regular <- mean(data.samp)
	prct.err.150.regular <- 100 * abs(mean(data.samp) - mean(simd1$sim1)) / mean(simd1$sim1)

	return(c(samp.mean.50.random, prct.err.50.random,
					 samp.mean.150.random, prct.err.150.random,
					 samp.mean.50.stratified, prct.err.50.stratified,
					 samp.mean.150.stratified, prct.err.150.stratified,
					 samp.mean.50.regular, prct.err.50.regular,
					 samp.mean.150.regular, prct.err.150.regular))
}

cleanup <- function(U, reps) {
	values1 <- c(1)
	values2 <- c(2)
	values3 <- c(3)
	values4 <- c(4)

	for(i in 1:(reps - 1)) {
		values1 <- c(values1, last(values1)+4)
		values2 <- c(values2, last(values2)+4)
		values3 <- c(values3, last(values3)+4)
		values4 <- c(values4, last(values4)+4)
	}

	sims1 <- as.data.frame(U) %>%
		select(all_of(values1))

	outputs1 <- tibble(mean = apply(sims1, 1, mean),
				 sd = apply(sims1, 1, sd))

	sims2 <- as.data.frame(U) %>%
		select(all_of(values2))

	outputs2 <- tibble(mean = apply(sims2, 1, mean),
										 sd = apply(sims2, 1, sd))

	sims3 <- as.data.frame(U) %>%
		select(all_of(values3))

	outputs3 <- tibble(mean = apply(sims3, 1, mean),
										 sd = apply(sims3, 1, sd))

	sims4 <- as.data.frame(U) %>%
		select(all_of(values4))

	outputs4 <- tibble(mean = apply(sims4, 1, mean),
										 sd = apply(sims4, 1, sd))

	outputs <- bind_cols(outputs1, outputs2, outputs3, outputs4)

	return(outputs)

}
```

```{r simulations, include=FALSE}
returns <- c("sample_mean*50.random", "percent_error*50.random",
						 "sample_mean*150.random", "percent_error*150.random",
						 "sample_mean*50.stratified", "percent_error*50.stratified",
						 "sample_mean*150.stratified", "percent_error*150.stratified",
						 "sample_mean*50.regular", "percent_error*50.regular",
						 "sample_mean*150.regular", "percent_error*150.regular")

reps <- 100

U <- replicate(reps, sapply(models, rand_samp))

outputs <- bind_cols(returns, cleanup(U, reps))

names(outputs) <- c("output", "mean1", "sd1", "mean2", "sd2", "mean3", "sd3", "mean4", "sd4")

outputs <- separate(outputs, col = "output", into = c("sample size", "type"), sep = "\\.")
outputs <- separate(outputs, col = "sample size", into = c("output", "sample size"), sep = "\\*")

outputs <- outputs %>%
	mutate(output = str_replace(output, pattern = "_", replacement = " "))

```

```{r models, echo=FALSE, message=FALSE, fig.align="center", fig.cap="Simulated models"}
grid.arrange(plot1, plot2, plot3, plot4)
```

\newpage

```{r sampling, echo=FALSE, message=FALSE, fig.align="center", fig.cap="Sampling approaches", fig.height=10}
grid.arrange(plot1.1, plot2.1, plot3.1,
						 plot1.2, plot2.2, plot3.2,
						 plot1.3, plot2.3, plot3.3,
						 plot1.4, plot2.4, plot3.4,
						 ncol = 2)
```

\newpage

```{r results, echo=FALSE, message=FALSE, fig.align="center"}
outputs %>% kable(caption = "Sampling results", digits = 3)
```

Table 1 shows the results across all sampling approaches, across all 4 models. Means (and standard deviations), 1 through 4, refer to models 1 through 4. Specifically, Model 1 is spatially random, Model 2 has a linear trend in the x direction, Model 3 has a Gaussian spatial dependence and Model 4 combines the linear trend and the Gaussian dependence. Means (and standard deviations), 1 through 4, represent the values calculated over 100 repetitions of the listed sampling approaches. 

In the case of simple random sampling, as the sample size increases from 50 to 150, Model 1 stops having the smallest sampling percent error (7.95%) and Model 4 overtakes (with 1.4% error). Overall, increased sample size reduces percent error across all 4 models. 

In the case of stratified random sampling, as the sample size increases from 50 to 150, Model 3 sampling retains the smallest overall percent error (3.7% with sample size 50 and 6.33% with sample size 150). Overall, increased sample size reduces percent error across almost all models, except for Model 3 where it increases. This indicates that this sampling approach might not be ideal for Model 3. 

In the case of systematic sampling, as the sample size increases from 50 to 150, Models 1,2 and 4 dramatically reduce their percent error while once again Model 3 sees an increase in percent error with a larger sample size. This makes Model 3 drop from having smallest percent error at sample size 50 to having the largest error at sample size 150. 

Overall, notwithstanding the effects of sample size, the best approach to sampling from the Model 1 distribution is systematic and for Models 2, 3, and 4 it is simple random sampling. This is based on sampling percent errors. Each error is actually the average error across a 100 simulations (with the standard deviations of these errors reported as well). 

\newpage

``` {r session, echo = FALSE}
sessionInfo()
```
